\bf Interactive Visualization of Hierarchically Structured Data

### 2017-07-12 reviewer_response.tex

We would like to thank the two reviewers and the associate editor for insightful
and constructive comments on our work. Below, we describe changes to the paper
text and software based on this feedback.

\subsection{Reviewer 1}

\textit{This small sampling of a large body of work points to the need for a
  cleaner conceptualization and clarifications of the contribution of the work
  presented in this paper relative to prior efforts.}

We have added material at the beginning of the literature review (Section 1.3) that
describes the larger body of work within which our study lives. We explain some of
the main ideas in the surveys you mentioned before proceeding to the more narrow
set of ideas that are directly adapted in our work. Further, we have provided a
bullet list in the introduction that explicitly describes the contribution of
this paper.

@@

\textit{My main concern regarding the presentation of the work lies in the third
  and fourth views - why are they presented as distinct? It would seem to me
  that all of the described functionality could be configured in that one view.
  In fact, I could imagine including the DOi in the same view as well - why are
  multiple views needed?}

The idea of combining these displays is interesting, and we have attempted to
implement a version which would highlight the series / nodes that lay in both
(1) the union of all treeboxes and (2) the intersection of all timeboxes.
However, we found this idea difficult to implement, for technical rather than
conceptual reasons -- when attempting to have both brush interactions for
querying and mouseover interactions for providing detail, the events conflict.
Indeed, we implemented the first and third recommendations described in
\href{https://github.com/d3/d3/issues/1604}{this D3 issue}, but in both cases we
found it necessary to introduce a lag between mousedown and the start of
brushing in order for the event to properly register. This lag created a drag on
responsiveness, and the value gained by unifying the displays did not seem worth
the cost of less fluid interactivity.

@@

\textit{Some detail about the scale of the tool would be useful. How many nodes
  can be included in the tree? How many time series, of what length? Rough
  estimates would help readers assess utility.}

We have added a discussion of scalability at the end of Section 1.4. We also
explicitly list the problem dimensions in Table 1. Also, all examples (and more)
are available as pre-compiled Rmarkdown vignettes linked from our package
webpage, so users can assess utility in this examples easily without having to
even install the package.

### 2017-07-12 treelapse.tex
update captions in all the microbiome examples

\begin{table}
\centering
\begin{tabular}{lll}
Data            & Number of timepoints & Number of nodes \\
Antibiotics     & 56                 & 386             \\
Preterm births  & 216                & 318             \\
Housing prices  & 254                & 944             \\
Bikesharing     & 24                 & 819             \\
Global patterns & 500                & 51             
\end{tabular}
\caption{Problem dimensions for each of the case studies. For problems with
  dimensions larger than that in the housing prices problem, we recommend an
  initial summarization or filtering step to prevent performance
  issues. \label{problem-scaling}}
\end{table}

@@

\caption{The two antibiotic time courses are readily apparent even when only
  highlighting the most abundant taxa. We have annotated the figure with its
  three main components.}\label{fig:antibioticoverview}
\end{figure}

@@

\caption{Introducing a second box into the timebox display identifies the
  Bacteroidaceae as a taxon only midly impacted by
  antibiotics.}\label{fig:antibioticbacteroidetes}
\end{figure}

@@

\caption{Zooming into the second antibiotic timecourse and highlighting the
  Lachnospiraceae and Ruminococcus, we see that the Ruminoccocus took more time
  to recover to pre-treatment levels.}\label{fig:antibioticfirmicutes}
\end{figure}

@@

\caption{By hovering over the Ruminoccocus branches, we see that there is a
  prolonged effect of the antibiotics time courses more or less uniformly across
  the lower taxonomic orders.}\label{fig:ruminococcus}
\end{figure}

@@

\caption{The subtree averages aggregation brings attention to the
  Verrucomicrobiae, which though only present as a few species, are each rather
  abundant. In particular, they seem to increase after the first antibiotic time
  course.}\label{fig:verrucomicrobiae}
\end{figure}

@@

\caption{The increased diversity among samples in CST 4 is represented by the
  relatively larger contribution of red edges to branches outside of the
  Firmicutes.}\label{fig:pretermcsts}
\end{figure}

@@

\caption{Zooming into the Lactobacillaceae family, we notice that the difference
  between the remaining four CSTs is related to which types of Lactobacillus are
  most prominent.}\label{fig:pretermcstslacto}
\end{figure}

@@

\caption{Samples with high levels of phyla other than Firmicutes appear to be
  related to preterm births.}\label{fig:pretermpreterm}
\end{figure}

>>>>

\begin{table}
\centering
\begin{tabular}{|l|l|l|}
  \hline
  Data            & Number of timepoints & Number of nodes \\
  \hline
  Antibiotics     & 56                 & 386             \\
  Preterm births  & 216                & 318             \\
  Housing prices  & 254                & 944             \\
  Bikesharing     & 24                 & 819             \\
  Global patterns & 500                & 51             \\
  \hline
\end{tabular}
\caption{Problem dimensions for each of the case studies. For problems with
  dimensions larger than that in the housing prices problem, we recommend an
  initial summarization or filtering step to prevent performance
  issues. \label{problem-scaling}}
\end{table}

@@

\caption{Here we display the primary timebox tree view of the antibiotics data
  set from \citep{dethlefsen2008}, annotated with the main components of the
  visualization. The tree at the top is a taxonomic tree of all the bacteria
  contained within the sample, and it is visually linked to the time series at
  the bottom: each node in the tree corresponds to a path among the time series.
  The selection brush is used to focus attention on the time series that go
  through it -- these are highlighted in blue -- and other brushes can be added
  using a button not displayed here. The pan-zoom widget at the top right is
  used to update the scales of the main time series display so that only
  particular time windows and $y$-axis ranges are visible. This view is the
  basis for all the timebox tree and treebox displays that appear
  below. \label{antibioticoverview}}
\end{figure}

@@

\caption{Introducing a second box into the timebox display identifies the
  Bacteroidetes as a taxon only midly impacted by antibiotics. The layout is
  identical to Figure 1, except two small brushes are placed over the time
  series between 10 and 20 days, and now only those time series and
  corresponding nodes in the tree are highlighted in blue. Further, the user has
  hovered over the top blue node in the tree, revealing the taxonomic identity
  of these series. Hence, brushing the time series and linking with the tree can
  be used to discover and characterize notable variation within the
  data.\label{fig:antibioticbacteroidetes}}
\end{figure}

@@

\caption{Zooming into the second antibiotic timecourse and highlighting the
  Lachnospiraceae and Ruminococcus, we see that the Ruminoccocus took more time
  to recover to pre-treatment levels. Here, the red lines and nodes are those
  that match the text search provided by the user in a search box just below the
  figure (not displayed here). Hovering the mouse over these lines provides
  their identities -- the top red line is Lachnospiraceae, and the bottom red
  line is Ruminoccocus. Note that the brush in the treebox display is located
  over the tree, rather than over the time series. In particular, the search box
  and interactive brushing can be combined to interrogate hypotheses of apriori
  interest. \label{fig:antibioticfirmicutes}}
\end{figure}

@@

\caption{By hovering over the Ruminoccocus branches, we see that there is a
  prolonged effect of the antibiotics time courses more or less uniformly across
  the lower taxonomic orders. The graphical elements are the same as before,
  except the user has searched for Ruminoccocus and species Unc095d3, which has
  the highest average abundance within this taxon. By displaying averages rather
  than sums, we see that the effect of antibiotics visible at higher taxonomic
  orders is not created by a single dominant species becoming less abundant, but
  rather the decline in populations across all descendant species. The same
  display applied to different data can yield different
  insights.}\label{fig:ruminococcus}
\end{figure}

@@

\caption{The subtree averages aggregation brings attention to the
  Verrucomicrobiae, which though only present as a few species, are each rather
  abundant. In particular, they seem to increase after the first antibiotic time
  course, which occurs between days 15 and 20. This view was generated by
  placing a brush over the branch on the far left, which has those nodes with
  the largest averages across all timepoints. The user's mouse is over the blue
  series, which brings up the associated taxonomic label. The determination of
  species whose abundances increase during antibiotics, which would require many
  hypothesis tests using a more standard approach, becomes quickly apparent
  via interactive visualization.}\label{fig:verrucomicrobiae}
\end{figure}

@@

\caption{The increased diversity among samples in community state type (CST) 4
  is represented by the relatively larger contribution of red edges to branches
  outside of the Firmicutes. This display shows the top of the DOI sankey
  visualization of the preterm birth data studied in
  \citep{digiulio2015temporal}. The root of the tree is the taxonomic kingdom
  Bacteria, its children are labeled according to their phylum names. Each color
  within a branch is associated with CST, and the width of the associated color
  corresponds to the average abundance of that taxonomic group among all samples
  belonging to the corresponding CST, as indicated by the legend at the left.
  Phyla are sorted from most to least abundant. This initial display of the DOI
  sankey provides a summary of overall abundances across taxonomic groups and
  CSTs, and suggests subtrees to navigate into, to extract more detailed
  abundance information.
}\label{fig:pretermcsts}
\end{figure}

@@

\caption{Zooming into the Lactobacillaceae family, we notice that the difference
  between the remaining four CSTs is related to which types of Lactobacillus are
  most prominent. The DOI sankey refocuses the tree around the last group that
  was clicked on, in this case showing more detail about order Lactobacillales
  and its descendants. The overview display can be recovered by navigating back
  up to higher-level taxa. Hence, it is possible to navigate between broad
  overview and detailed displays in a way that facilitates interpretation of
  results from statistical analysis.}\label{fig:pretermcstslacto}
\end{figure}

@@

\caption{Samples with high levels of phyla other than Firmicutes appear to be
  related to preterm births. Here, we again display the preterm birth data with
  a DOI sankey, but instead of grouping samples according to statistically
  generated CSTs, we directly assign samples to preterm vs. not preterm
  according to whether the mother eventually had a preterm birth. These new
  states are visible in the updated legend. Through interactivity, it becomes
  possible to develop meaningful visual summaries even before calculating formal
  statistical ones.}
}\label{fig:pretermpreterm}
\end{figure}
